Unsupervised embeddings of organizations capture latent structure of scientific mobility

### 2020-03-03 paper/current/main.tex
revise another para

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.7\textwidth]{\figdir/mobility_example_combined.pdf}
    \caption{
        \textbf{Co-affiliation trajectories represented as sentences}
        \textbf{a.} 
        An author published five papers across five time periods, with only one affiliation listed in the byline of each paper. 
        A unique identifier is assigned to each organization, and these are concatenated into a mobility sentence. 
        \textbf{b.}
        In a more complex case, an author lists several affiliations on the byline of papers in $t_{2}$, $t_{3}$, and $t_{4}$. Here, organization ids are still ordered by year, however order within a single year is random.      
    }
    \label{fig:methods:mobility_sentence}
\end{figure}

@@

The \SGNS{} model learns an effective distance that explains and predicts scientific mobility better than alternatives. 
Embedding distance explains more of the expected flow between pairs of organizations, with correlation ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) over double that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
Embedding distance also explains more of the variance of domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and international mobility flows (Fig. ~\ref{fig:supp:gravity_correlation}d) than does geographic distance. 
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance globally (Fig.~\ref{fig:gravity_performance}c-d) and for domestic (Fig.~\ref{fig:supp:gravity_correlation}e-f) and international mobility(Fig.~\ref{fig:supp:gravity_correlation}g-h).
Similarly, embedding distance outperforms distances derived from calculating personalized-page rank scores on the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}). 
Geographic distance alone is a poor tool for understanding international scientific mobility---rather, by learning a distance directly from the data, it is possible to better explain, and better predict flux between organizations.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance outperforms and is more robust than geographic distance.}
		\textbf{a.} embedding distance (cosine distance between organization vectors) explains more of the expected flux between organization pairs than does geographic distance (\textbf{b}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		\textbf{c-d.} Predictions  of flux between organization pairs made using embedding distance out-perform those made using geographic distance. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		\textbf{e-f.} Predictions made using model parameters learned from subset of same-country organization pairs. 
		Embedding distance outperforms those made for geographic distance.
		\textbf{g-h.} Embedding distance also outperforms geographic distance when predicting flux using model parameters estimated from only pairs of organizations in different countries. 
		Predictions are not meaningfully different if estimates are learned on all pairs of organizations (Fig. \ref{fig:supp:gravity_prediction_subsets}).
		These predictions are robust whether the model is fit on the global data, or on the domestic (Fig.~\ref{fig:gravity_performance}e-h) or international Fig.~\ref{fig:supp:gravity_prediction_subsets}) subsets .
	}
	\label{fig:gravity_performance}
\end{figure}

@@

Why does the SNGS model, a tool designed for natural language processing, produce such a reasonable distance that models gravity relationships between organizations?
We find evidence of a theoretical connection between the two models, one that makes the skip-gram model applicable to many other forms of mobility trajectory.

Let $f_{ij}$ represents the total \textbf{flux}, the number of people who moved between organizations $i$ and $j$. $M_{i}$ and $M_{j}$ represent the size of an organization. According to our gravity model on the embedding space,

\begin{equation}
    \label{eq:gravity_model}
    \frac{f_{ij}}{M_{i} * M_{j}} \propto \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    \label{eq:gravity_model_approx}
    \frac{f_{ij}}{\mu_i * \mu_j} \propto \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    \label{eq:gravity_relationship}
    \frac{f_{ij}}{\mu_i * \mu_j} = \frac{\mu_i * p(j|i)}{\mu_i * \mu_j} =  \frac{p(j|i)}{\mu_j} \propto \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    \label{eq:conditional probability}
      p(j|i) \propto {\mu_j}* \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    p(j|i) \propto \frac{{\mu_j}* \exp(\cos(v_i,v_j)}{\sum_{k\in\mathpzc{N}_i} {\mu_k}* \exp(\cos(v_i,v_k))}
\end{equation}

>>>>

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.7\textwidth]{\figdir/mobility_example_combined.pdf}
    \caption{
        \textbf{Co-affiliation trajectories represented as sentences}
        \textbf{a.} 
        An author published five papers across five time periods, with only one affiliation listed in the byline of each paper. 
        To each organization, a unique identifier is assigned, and they are assembled into a mobility sentence. 
        \textbf{b.}
        In a more complex case, an author lists several affiliations on the byline of papers in $t_{2}$, $t_{3}$, and $t_{4}$. Here, organization ids are still ordered by year, however order within a single year is random.      
    }
    \label{fig:methods:mobility_sentence}
\end{figure}

@@

The effective distance  estimated by our embedding model explains and predicts scientific mobility best. 
Embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds for the variance of domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and international mobility flows (Fig.~\ref{fig:supp:gravity_correlation}d). 
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance globally (Fig.~\ref{fig:gravity_performance}c--d) and for domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and international mobility(Fig.~\ref{fig:supp:gravity_correlation}g--h).
Similarly, embedding ``diffusion'' distance outperforms distances derived from calculating personalized-page rank scores on the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}). 
Geographic distance alone is a poor tool for understanding international scientific mobility---rather, by learning a distance directly from the data, it is possible to better explain, and better predict flux between organizations.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance outperforms and is more robust than geographic distance.}
		\textbf{a.} embedding distance (cosine distance between organization vectors) explains more of the expected flux between organization pairs than does geographic distance (\textbf{b}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		\textbf{c-d.} Predictions  of flux between organization pairs made using embedding distance outperform those made using geographic distance. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		\textbf{e-f.} Predictions made using model parameters learned from subset of same-country organization pairs. 
		Embedding distance outperforms those made for geographic distance.
		\textbf{g-h.} Embedding distance also outperforms geographic distance when predicting flux using model parameters estimated from only pairs of organizations in different countries. 
		Predictions are not meaningfully different if estimates are learned on all pairs of organizations (Fig. \ref{fig:supp:gravity_prediction_subsets}).
		These predictions are robust whether the model is fit on the global data, or on the domestic (Fig.~\ref{fig:gravity_performance}e-h) or international Fig.~\ref{fig:supp:gravity_prediction_subsets}) subsets .
	}
	\label{fig:gravity_performance}
\end{figure}

@@

Why does the word embedding model, a tool designed for natural language processing, produce such a reasonable distance that models gravity relationships between organizations?
We find evidence of a theoretical connection between the two models, one that makes the skip-gram model applicable to many other forms of mobility trajectory.

Let $t_{ij}$ represents the total \textbf{flux}, the number of people who moved between organizations $i$ and $j$. $M_{i}$ and $M_{j}$ represent the size of an organization. According to our gravity model on the embedding space,

\begin{equation}
    \label{eq:gravity_model}
    \frac{f_{ij}}{M_{i} M_{j}} \propto \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    \label{eq:gravity_model_approx}
    \frac{f_{ij}}{\mu_i  \mu_j} \propto \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    \label{eq:gravity_relationship}
    \frac{f_{ij}}{\mu_i  \mu_j} = \frac{\mu_i  p(j|i)}{\mu_i\mu_j} =  \frac{p(j|i)}{\mu_j} \propto \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    \label{eq:conditional probability}
      p(j|i) \propto {\mu_j} \exp(\cos(v_i,v_j))
\end{equation}

@@

\begin{equation}
    p(j|i) \propto \frac{{\mu_j}\exp(\cos(v_i,v_j)}{\sum_{k\in\mathpzc{N}_i} {\mu_k}\exp(\cos(v_i,v_k))}
\end{equation}
